Place of death in NRS COVID-19 deaths

### 2021-02-12 covid_nrs_place_of_death.Rmd
fixed & included map data & my first map plot - ratio of home death 2020 to historical mean. need to configure size of text to plot etc.

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(maps)
```

@@

TODO: figure out which boundaries in Scotland correspond to council areas!

~~To get council areas in Scotland we need NUTS 2 level map data~~, and the most appropriate level of detail to use is likely the BUC - Ultra Generalised Clipped

https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom

https://geoportal.statistics.gov.uk/datasets/nuts-level-2-january-2018-ultra-generalised-clipped-boundaries-in-the-united-kingdom?geometry=-12.122%2C55.353%2C4.566%2C57.479

health board boundaries April 2019, BUC
https://opendata.arcgis.com/datasets/def40bdc98a9457aa108eb3a5fb052b1_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D

getting error trying to use shapefile with ggplot:
> Error in CPL_transform(x, crs, aoi, pipeline, reverse) : 
  crs not found: is it missing?

explicitly setting crs on shapefile fails,

version fo sf 0.9-7:
> Linking to GEOS 3.8.0, GDAL 3.0.4, PROJ 6.3.1

```{r}
# TODO: learn how to use maps etc.
zip_file <- tempfile()
unzipped <- tempfile()

## Scottish LAU 2011 level 1
# curl_download(url = "https://borders.ukdataservice.ac.uk/ukborders/easy_download/prebuilt/shape/Scotland_laulevel1_2011.zip", destfile = zip_file, quiet = FALSE)  # download file from ONS
# name <- list.files(unzipped, pattern=".shp")
# shapefile <- st_read(file.path(unzipped, name))

## below fails, saying cannot allocate vector of size XX MB/GB
# curl_download(url = "https://opendata.arcgis.com/datasets/7f83b82ef6ce46d3a5635d371e8a3e7c_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D", destfile = zip_file, quiet = FALSE)  # download file from ONS

download_and_import_shapefile <- function(url) {
  zip_file <- tempfile()
  unzipped <- tempfile()
  downloaded_zip <- curl_download(url = url, destfile = zip_file, quiet = FALSE) 
  unzip(zipfile = zip_file, exdir = unzipped)
  name <- list.files(unzipped, pattern=".shp$")
  shapefile <- st_read(file.path(unzipped, name))
  return(shapefile)
}

health_boards_shapefile_url <- "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip"
la_shapefile_url <- "https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.zip?outSR=%7B%22latestWkid%22%3A27700%2C%22wkid%22%3A27700%7D"

shapefile_hb <- download_and_import_shapefile(url = health_boards_shapefile_url)
shapefile_la <- download_and_import_shapefile(url = la_shapefile_url)

# shapefile %>%
#   ggplot(aes(geometry = geometry)) +
#   geom_sf()

# shapefile_la %>%
#   inner_join(merged_deaths_la, by = c("lad19nm"="ref_area"))

# shapefile_hb %>%
#   inner_join(merged_deaths_hb, by = c("HBName"="ref_area"))

# TODO: summarise data by LA/HB so it can be shown on a map
# otherwise it loads forever because we're re-plotting the same regional boundary for each row!!

merged_deaths_la %>%
  filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
  left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
  ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
  geom_sf()

# ggplot(data = shapefile_hb, aes(geometry=geometry)) + geom_sf()

# merged_deaths_hb %>%
#   left_join(shapefile_hb,by=c("ref_area"="HBName")) %>%
#   ggplot(aes(fill = deaths_mean, geometry = geometry)) +
#   geom_sf()

# ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
#   geom_polygon() +
#   coord_map()
```

@@

historical_average_sex <-
  past_deaths_sex %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
```

@@

```{r}
## compute rolling average

merged_data_overall <-
  weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  inner_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death", "week_number", "sex", "age"))

merged_deaths_hb <-
  # weekly_deaths_hb %>%
  rolling_deaths_hb %>%  # use data with rolling movingincluded
  inner_join(historical_average_hb, by = c("week_number","ref_area","place_of_death"))

merged_deaths_la <-
  # weekly_deaths_la %>%
  rolling_deaths_la %>%  # use data with rolling movingincluded
  inner_join(historical_average_la, by = c("week_number","ref_area","place_of_death"))

merged_data_sex_age <-
  # weekly_deaths_sex_age %>%
  rolling_deaths_sex_age %>%  # use data with rolling movingincluded
  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))

merged_data_sex <-
  rolling_deaths_sex %>%
  inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))

merged_data_age <-
  rolling_deaths_age %>%
  inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))
```

@@

@@

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")

annotations_figure_place_of_death <-
# datapasta::tribble_paste(merged_data_overall %>% select(place_of_death))
  tibble::tribble(
    ~place_of_death, ~x, ~y, ~label,
                 "Hospital", 20L, 850, "Shaded area shows range of 2015-2019 deaths; dashed line shows weekly average",
                 "Hospital", 25L, 250, "Hospital deaths dipped below historical minimum and remained low until Winter.",
                "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
                     "Home & other non-institution", 35L, 500, "The gap between historical high and both types of death at home was high throughout."
                  ) %>% 
  mutate(
    place_of_death = factor(place_of_death, levels = levels(merged_data_overall$place_of_death)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
    label = str_wrap(label, width = 40)
    )

figure_place_of_death <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  geom_text(data = annotations_figure_place_of_death, aes(x=x,y=y, label=label)) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
figure_deaths_at_home <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_2020 <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(year == 2020) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       # subtitle = date_range_text,
       subtitle = "Data from 2020 shown, starting with w/c 30 Dec 2019.",
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_data_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_age %>% filter(year==2021) %>% .$week_number)} in 2021.")

figure_deaths_at_home_by_age <-
  merged_data_age %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(!age %in% c('0','1-14')) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  # facet_wrap(~age, scales = "free_y") +
  facet_wrap(~age) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text,removed_under_15_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_by_sex <-
  merged_data_sex %>%
  mutate(sex = if_else(sex=="F", "Female", "Male")) %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~sex) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r}
# TODO: add yearly range for all deaths
# TODO: split data by covid-underlying cause of desath & covid-factor in death
merged_data_overall %>%
  filter(sex=="all", age=="all") %>%
  filter(year==2020) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  group_by(ref_area, year) %>%
  summarise(
    deaths_all_causes = sum(deaths_all_causes, na.rm=TRUE),
    deaths_covid_related = sum(deaths_covid_related, na.rm=TRUE),
    deaths_non_covid = sum(deaths_non_covid, na.rm=TRUE)
  )
```

>>>>

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
```

@@

I previously downloaded shapefiles for:

* Local Authority Districts (LAD) for 2020 (BUC-level resolution): https://geoportal.statistics.gov.uk/datasets/local-authority-districts-may-2020-boundaries-uk-buc
* Health Boards https://data.gov.uk/dataset/27d0fe5f-79bb-4116-aec9-a8e565ff756a/nhs-health-boards
  - Note: this is a fairly high resolution, don't need this much detail!
  - I found a discussion of this on [a github covid-related project](https://github.com/tomwhite/covid-19-uk-data/issues/18#issuecomment-622353841), and user **robchallen** produced a zip file containing [UK-wide health board boundaries](https://github.com/tomwhite/covid-19-uk-data/files/4563933/UK_covid_reporting_regions.zip)

```{r}
# shapefile_hb <- st_read(dsn = "./map_data/hb")  # too detailed!
shapefile_hb_uk_wide <- st_read(dsn = "./map_data/hb_uk_wide/UK_covid_reporting_regions")
shapefile_la <- st_read(dsn = "./map_data/la")

# merged_deaths_la %>%
#   filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
#   left_join(shapefile_la,by=c("ref_area"="lad19nm")) %>%
#   ggplot(aes(fill = deaths_all_causes, geometry = geometry, group=ref_area)) +
#   geom_sf()
# 
# merged_deaths_hb %>%
#   filter(week_number==53 & place_of_death=="Home & other non-institution") %>%
#   left_join(shapefile_hb_uk_wide,by=c("ref_area"="name")) %>%
#   ggplot(aes(fill = deaths_all_causes, geometry = geometry)) +
#   geom_sf() +
#   theme(aspect.ratio = 1.3)

# ggplot(data = shapefile_hb_uk_wide, aes(geometry=geometry)) + geom_sf()

## simple UK map
# ggplot(data = map_data(map = "world", region = "UK"), aes(x=long, y = lat, group = group)) +
#   geom_polygon() +
#   coord_map()
```

@@

historical_average_sex <-
  past_deaths_sex %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

compute_historical_annual_average_deaths <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, !!!grouping_vars) %>%
    summarise(  # calculate annual totals first...
    number_of_deaths = sum(number_of_deaths), 
    .groups = "keep"
  ) %>%
  ungroup(year) %>%  # remove year as grouping to average over year now
  summarise(
    annual_deaths_mean = mean(number_of_deaths),
    annual_deaths_min = min(number_of_deaths),
    annual_deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
}

historical_annual_average_hb <-
  past_deaths_hb %>%
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_la <-
  past_deaths_la %>%
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_sex_age <-
  past_deaths_sex_age %>%
  compute_historical_annual_average_deaths(sex, age)

historical_annual_average_sex <-
  past_deaths_sex %>%
  compute_historical_annual_average_deaths(sex)

historical_annual_average_age <-
  past_deaths_age %>%
  compute_historical_annual_average_deaths(age)
```

@@

```{r}
## compute rolling average
merged_deaths_overall <-
  weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  inner_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death", "week_number", "sex", "age"))

merged_deaths_hb <-
  rolling_deaths_hb %>%  # use data with rolling movingincluded
  inner_join(historical_average_hb, by = c("week_number","ref_area","place_of_death"))

merged_deaths_la <-
  rolling_deaths_la %>%  # use data with rolling movingincluded
  inner_join(historical_average_la, by = c("week_number","ref_area","place_of_death"))

merged_deaths_sex_age <-
  rolling_deaths_sex_age %>%  # use data with rolling movingincluded
  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))

merged_deaths_sex <-
  rolling_deaths_sex %>%
  inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))

merged_deaths_age <-
  rolling_deaths_age %>%
  inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))
```

# Summarise annual deaths for 2020 & 2021

For visualising data using maps, we need to compute data for the entire year. Note: the non-covid and covid-related deaths in 2020 don't add up to all cause deaths due to the covid status only starting in week 12!

```{r}
compute_total_deaths_in_year <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, !!!grouping_vars) %>%
    summarise(across(.cols = c(deaths_all_causes, deaths_covid_related, deaths_non_covid), .fns = ~sum(.x,na.rm=TRUE)), .groups = "drop") %>%
    mutate(deaths_all_causes_since_covid = deaths_covid_related + deaths_non_covid)
}

compute_ratio_of_annual_deaths_to_historical <- function(data_tbl) {
  data_tbl %>%
    mutate(
      ratio_annual_deaths_to_historical_mean = deaths_all_causes / annual_deaths_mean,
      ratio_annual_deaths_to_historical_min = deaths_all_causes / annual_deaths_min,
      ratio_annual_deaths_to_historical_max = deaths_all_causes / annual_deaths_max
    )
}

total_deaths_la <-
  merged_deaths_la %>%
  compute_total_deaths_in_year(ref_area) %>%
  left_join(historical_annual_average_la, by = c("place_of_death", "ref_area")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_hb <-
  merged_deaths_hb %>%
  compute_total_deaths_in_year(ref_area) %>%
  left_join(historical_annual_average_hb, by = c("place_of_death", "ref_area")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_sex <-
  merged_deaths_sex %>%
  compute_total_deaths_in_year(sex) %>%
  left_join(historical_annual_average_sex, by = c("place_of_death", "sex")) %>%
  compute_ratio_of_annual_deaths_to_historical

total_deaths_age <-
  merged_deaths_age %>%
  compute_total_deaths_in_year(age) %>%
  left_join(historical_annual_average_age, by = c("place_of_death", "age")) %>%
  compute_ratio_of_annual_deaths_to_historical
```

@@

only_2020_text <- glue::glue("Data from 2020 included, starting with w/c {format(min(merged_deaths_age$date_w_c))}.")

@@

## Total deaths in 2020 by age group & place of death

```{r, fig.width=12, fig.height=7, warning=FALSE}
bar_and_error_bars_2020_text <- "Bars are 2020 all-cause deaths, point and bars shows annual mean and range 2015-2019"
excluded_other_place_2020 <- glue::glue("Excluding records with 'Other' place of death, N={total_deaths_age %>% filter(year==2020 & place_of_death=='Other') %>% .$deaths_all_causes %>% sum}")

light_shaded_area_shows_difference <- "Lighter area in the top of bar shows difference between 2020 deaths and historical maximum 2015-2019"

# TODO: make the distinction max vs 2020 figures clearer! maybe annotate?

figure_annual_deaths_by_age_group_2020 <-
  total_deaths_age %>% 
  filter(year==2020) %>%
  filter(place_of_death != "Other") %>%
  ggplot(aes(x = age, y = deaths_all_causes)) +
  # geom_col(fill = "#4477AA", alpha = 0.5, colour = "grey10") +
  geom_col(fill = "grey20", colour = "grey10", alpha = 0.5) +
  geom_col(aes(x=age, y=annual_deaths_max), fill = "#4477AA", colour = "grey10", alpha = 0.5) +  # historical maximum over the actual figure
  geom_point(aes(x=age, y=annual_deaths_mean)) +
  geom_errorbar(aes(x = age, ymin = annual_deaths_min, ymax = annual_deaths_max)) +
  facet_wrap(~place_of_death) +
  scale_y_continuous(labels = function(x) format(x,big.mark=",")) +
  theme_minimal() +
  labs(x = "Age",
       y = "Deaths (N)",
       title = "Deaths in 2020 by age group & place of death",
       subtitle = paste0(c(only_2020_text,light_shaded_area_shows_difference),collapse="\n"),
       caption = paste0(c(excluded_other_place_2020,bar_and_error_bars_2020_text,death_registrations_text,source_nrs_text), collapse="\n")
  )

walk(
  .x = c(".pdf",".svg",".png"),
  .f = ~ggsave(filename = paste0("./outputs/figure_annual_deaths_by_age_group_2020",.x), plot = figure_annual_deaths_by_age_group_2020, width = 12, height = 7, dpi = 300)
)

figure_annual_deaths_by_age_group_2020
```

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")

annotations_figure_place_of_death <-
# datapasta::tribble_paste(merged_deaths_overall %>% select(place_of_death))
  tibble::tribble(
    ~place_of_death, ~x, ~y, ~label,
                 "Hospital", 20L, 850, "Shaded area shows range of 2015-2019 deaths; dashed line shows weekly average",
                 "Hospital", 25L, 250, "Hospital deaths dipped below historical minimum and remained low until Winter.",
                "Care home", 35L, 500, "After initial peak, deaths returned to historical range; non-Covid deaths remained low.",
                     "Home & other non-institution", 35L, 500, "The gap between historical high and both types of death at home was high throughout."
                  ) %>% 
  mutate(
    place_of_death = factor(place_of_death, levels = levels(merged_deaths_overall$place_of_death)),  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!
    label = str_wrap(label, width = 40)
    )

figure_place_of_death <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  geom_text(data = annotations_figure_place_of_death, aes(x=x,y=y, label=label)) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
figure_deaths_at_home <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_2020 <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(year == 2020) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       # subtitle = date_range_text,
       subtitle = "Data from 2020 shown, starting with w/c 30 Dec 2019.",
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
removed_under_15_text <- glue::glue("Excluding individuals aged <15 at death, total N={merged_deaths_age %>% filter(age %in% c('0','1-14') & place_of_death == 'Home & other non-institution') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_deaths_age %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_deaths_age %>% filter(year==2021) %>% .$week_number)} in 2021.")

figure_deaths_at_home_by_age <-
  merged_deaths_age %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  filter(!age %in% c('0','1-14')) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  # facet_wrap(~age, scales = "free_y") +
  facet_wrap(~age) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text,removed_under_15_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r, fig.width=12, fig.height=9, warning=FALSE}
figure_deaths_at_home_by_sex <-
  merged_deaths_sex %>%
  mutate(sex = if_else(sex=="F", "Female", "Male")) %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~sex) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r}
# TODO: add yearly range for all deaths
# TODO: split data by covid-underlying cause of desath & covid-factor in death
merged_deaths_overall %>%
  filter(sex=="all", age=="all") %>%
  filter(year==2020) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  group_by(ref_area, year) %>%
  summarise(
    deaths_all_causes = sum(deaths_all_causes, na.rm=TRUE),
    deaths_covid_related = sum(deaths_covid_related, na.rm=TRUE),
    deaths_non_covid = sum(deaths_non_covid, na.rm=TRUE)
  )
```

# Map-based visualisations

## Ratio of 2020 home deaths to historical home deaths by LA
